## Exploratory Data Analysis

Helper functions to truncate data
```{r functions.eda}
#' winsorizes/truncates values for a given vector and quantile ranges
#'
#' @param vector vector with numerical values
#' @param lower_quantile_boundary lower quantile boundary
#' @param upper_quantile_boundary upper quantile boundary
#'
#' @return winsorized vector with min(vector) = lower_quantile and
#'         max(vector) = upper_quantile
winsorize <- function(vector, lower_quantile_boundary = 0.05,
                      upper_quantile_boundary = 0.95) {
  if (lower_quantile_boundary > upper_quantile_boundary) {
    warning("Lower quantile boundary is greater than upper quantile boundary.")
  }

  lower_quantile <- quantile(vector, lower_quantile_boundary, na.rm = TRUE)
  upper_quantile <- quantile(vector, upper_quantile_boundary, na.rm = TRUE)

  vector[!is.na(vector) & vector < lower_quantile] <- lower_quantile
  vector[!is.na(vector) & vector > upper_quantile] <- upper_quantile

  return(vector)
}

#' winsorizes/truncate columns of a given tibble
#'
#' @param data tibble with data
#' @param feature_identifier vector with strings of column names of tibble
#'        which should be winsorized
#' @param lower_quantile_boundary lower quantile boundary
#' @param upper_quantile_boundary upper quantile boundary
#'
#' @return tibble of the same structure as given with winsorized columns
data.prepare.winsorize <- function(data, feature_identifier,
                                   lower_quantile_boundary = 0.05,
                                   upper_quantile_boundary = 0.95) {
  data <- data %>%
    ungroup() %>%
    mutate(across(all_of(feature_identifier),
      .fns = ~ winsorize(.x, lower_quantile_boundary, upper_quantile_boundary)
    ))
  return(data)
}
```

### Correlation analysis
```{r, correlation_analysis}
correlation.info <- function(data, feature_identifier) {
  correlation.matrix <- data %>%
    select(all_of(feature_identifier)) %>%
    cor(x = , use = "pairwise.complete.obs", method = "pearson") %>%
    as.data.frame() %>%
    rownames_to_column(var = "rowname")

  correlation.plot <- correlation.matrix %>%
    pivot_longer(
      cols = -rowname,
      names_to = "rowname1",
      values_to = "correlation"
    ) %>%
    ggplot(aes(rowname, rowname1, fill = correlation)) +
    geom_tile() +
    scale_fill_gradient2(
      low = "blue", high = "red", mid = "white",
      midpoint = 0, space = "Lab",
      name = "Pearson\nCorrelation"
    ) +
    labs(
      title = "Correlation Matrix",
      x = "",
      y = ""
    ) +
    theme(
      aspect.ratio = 1,
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
    )

  return(list(matrix = correlation.matrix, plot = correlation.plot))
}

correlation.info.initial <- correlation.info(com, com.features.fin$identifier)
correlation.info.initial$plot
```

The majority of features is not correlated with a Pearson's correlation coefficient of around 0. The operative profit
margin ratios `opmad` and `opmbd` as well as the liquidity ratios `curr_ratio` and `quick_ratio` are perfectly positively correlated. The correlation coefficient between the debt ratios `debt_assets` and `debt_at` is 0.76. The remaining correlation coefficients are all lower than 0.75.  Therefore, the next chunk searches all pairwise correlations and removes one of the pairs if its absolute value is higher than the `CORRELATION_THRESHOLD`.

```{r correlated_features_detect_removals}
# winsorize features
com.winsorized <- data.prepare.winsorize(com, com.features.fin$identifier,
  lower_quantile_boundary = 0.05,
  upper_quantile_boundary = 0.95
)

n_features <- length(com.features.fin$identifier)
feature_indicators_to_remove <- character(0)
# check features mutually if they meet CORRELATION_THRESHOLD and remove one of them
for (i in 1:(n_features - 1)) {
  feature_identifier_i <- com.features.fin$identifier[i]
  for (j in (i + 1):n_features) {
    feature_identifier_j <- com.features.fin$identifier[j]
    if (abs(correlation.info.initial$matrix[i, j + 1]) >= CORRELATION_THRESHOLD) {
      # scatter plot of correlated features > truncation tbd
      print(ggplot(com.winsorized, aes_string(x = feature_identifier_i, y = feature_identifier_j)) +
        geom_point() +
        geom_smooth() +
        labs(title = paste("Scatterplot of", feature_identifier_i, "vs", feature_identifier_j)))

      print(paste0(
        "Correlation between features ", feature_identifier_i,
        " and ", feature_identifier_j, " is ",
        formatC(correlation.info.initial$matrix[i, j + 1], digits = 4),
        ". Therefore, feature ", feature_identifier_j, " will be removed in further analyses."
      ))

      feature_indicators_to_remove <- c(feature_indicators_to_remove, feature_identifier_j)
    }
  }
}
```
The two scatterplots of the highly correlated (winsorized) features `opmad` and `opmbd` as well as `curr_ratio` and `quick_ratio` confirm linear relationship between them. The scatterplot between `debt_assets` and `debt_at` shows non-linear structures for higher values.

### Features and delisting frequency
The following histograms show for each feature in the data set the number of data rows that fall in a specific range in gray (x-axis). The blue bars represent the number of financial ratio rows in a specific range that belong to a company which delisted in the subsequent year (gap of three month included). For example, if a company was delisted in April 2002, then the financial ratios from January 2001 to January 2002 would be attributed to the blue bar. The ratio between the blue and gray bar, named relative delisting frequency, is displayed by the orange points and a corresponding smoothed line (method `LOESS` with tri-cubic kernel). The right y-axis is the scale for the relative frequency values. The feature values are winsorized at the 5% and 95% quantile.
```{r, features_distribution_delisting_frequency.function}
# Function to scale secondary axis
scale_function <- function(x, scale, shift) {
  return(x * scale + shift)
}

# Function to scale secondary variable values
inv_scale_function <- function(x, scale, shift) {
  return((x + shift) / scale)
}

#' creates a histogram of features vs delisting frequency
#'
#' @param data tibble containing financial ratios for companies with a column
#'             named feature_identifier and years_to_delisting
#' @param feature_identifier technical identifier of feature of interest
#' @param feature_description readable name of feature for title
#' @param marginal.data vector of predicted delistings
#' @param marginal.legend legend for marginal plots
#'
#' @return histogram of fincancial ratio features with up to two smoothed
#'         relative delisting frequencies
hist_delisting_frequency <- function(
    data,
    feature_identifier,
    feature_description,
    marginal.data = NULL, marginal.legend = "Modeled delisting frequency",
    marginal.data2 = NULL, marginal.legend2 = "") {
  feature <- data[[feature_identifier]] # first column
  years_to_delisting <- data$years_to_delisting # first column
  max_x <- max(feature, na.rm = TRUE)
  min_x <- min(feature, na.rm = TRUE)
  n_bins <- 30
  bin_boundaries <- seq(min_x, max_x, length.out = n_bins + 1)
  bin_width <- bin_boundaries[2] - bin_boundaries[1]

  plot.data <- tibble(
    feature = feature,
    years_to_delisting = years_to_delisting,
    feature_bin_index = findInterval(feature, bin_boundaries),
    bin_centers = bin_boundaries[1] + bin_width * (feature_bin_index - 1.5)
  )

  # compute points
  if (!is.null(marginal.data)) {
    marginal.data <- plot.data %>%
      mutate(pred = marginal.data) %>%
      group_by(bin_centers) %>%
      summarize(relative_frequency_delisting_pred = mean(pred)) %>%
      select(x = bin_centers, yhat = relative_frequency_delisting_pred)
  }

  # compute points
  if (!is.null(marginal.data2)) {
    marginal.data2 <- plot.data %>%
      mutate(pred = marginal.data2) %>%
      group_by(bin_centers) %>%
      summarize(relative_frequency_delisting_pred = mean(pred)) %>%
      select(x = bin_centers, yhat = relative_frequency_delisting_pred)
  }

  # max of first y axis
  max_y_first <- plot.data %>%
    group_by(feature_bin_index) %>%
    summarize(frequency = n()) %>%
    select(frequency) %>%
    max()
  max_y_second <- 0.3 # max of second y axis
  min_y_first <- 0 # min of first y axis
  min_y_second <- 0 # min of second y axis

  # scale and shift variables calculated based on desired mins and maxes
  scale <- (max_y_second - min_y_second) / (max_y_first - min_y_first)
  shift <- min_y_first - min_y_second

  com.frequency_delisting <- plot.data %>%
    mutate(delisting_indicator = determine_delisting_property(years_to_delisting)) %>%
    group_by(bin_centers) %>%
    summarize(
      frequency = n(),
      frequency_delisting = sum(delisting_indicator)
    ) %>%
    mutate(relative_frequency_delisting = frequency_delisting / frequency)

  plot <- ggplot(plot.data) +
    geom_histogram(
      aes(x = feature, color = "Absolute frequency", fill = "Absolute frequency"),
      binwidth = bin_width
    ) +
    geom_histogram(
      aes(
        x = feature * ifelse(determine_delisting_property(years_to_delisting), 1, NA),
        color = "Absolute delisting frequency", fill = "Absolute delisting frequency"
      ),
      binwidth = bin_width
    ) +
    geom_point(
      data = com.frequency_delisting,
      mapping = aes(
        x = bin_centers, y = inv_scale_function(relative_frequency_delisting, scale, shift),
        color = "Relative delisting frequency"
      )
    )

  if (feature_identifier == "pe_exi") {
    plot <- plot + geom_smooth(
      data = com.frequency_delisting %>% filter(bin_centers <= 0),
      method = "loess",
      mapping = aes(
        x = bin_centers, y = inv_scale_function(relative_frequency_delisting, scale, shift),
        color = "Relative delisting frequency"
      )
    ) +
      geom_smooth(
        data = com.frequency_delisting %>% filter(bin_centers > 0),
        mapping = aes(
          x = bin_centers, y = inv_scale_function(relative_frequency_delisting, scale, shift),
          color = "Relative delisting frequency"
        )
      )
    if (!is.null(marginal.data)) {
      plot <- plot + geom_point(
        data = marginal.data %>% filter(x <= 0),
        mapping = aes(
          x = x, y = inv_scale_function(yhat, scale, shift),
          color = marginal.legend
        )
      ) +
        geom_smooth(
          data = marginal.data %>% filter(x <= 0),
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend
          ), linetype = "dashed"
        ) +
        geom_point(
          data = marginal.data %>% filter(x > 0),
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend
          )
        ) +
        geom_smooth(
          data = marginal.data %>% filter(x > 0),
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend
          ), linetype = "dashed"
        )
    }
    if (!is.null(marginal.data2)) {
      plot <- plot + geom_point(
        data = marginal.data2 %>% filter(x <= 0),
        mapping = aes(
          x = x, y = inv_scale_function(yhat, scale, shift),
          color = marginal.legend2
        )
      ) +
        geom_smooth(
          data = marginal.data2 %>% filter(x <= 0),
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend2
          ), linetype = "dotted"
        ) +
        geom_point(
          data = marginal.data2 %>% filter(x > 0),
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend2
          )
        ) +
        geom_smooth(
          data = marginal.data2 %>% filter(x > 0),
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend2
          ), linetype = "dotted"
        )
    }
  } else {
    plot <- plot + geom_smooth(
      data = com.frequency_delisting,
      mapping = aes(
        x = bin_centers, y = inv_scale_function(relative_frequency_delisting, scale, shift),
        color = "Relative delisting frequency"
      )
    )

    if (!is.null(marginal.data)) {
      plot <- plot + geom_point(
        data = marginal.data,
        mapping = aes(
          x = x, y = inv_scale_function(yhat, scale, shift),
          color = marginal.legend
        )
      ) +
        geom_smooth(
          data = marginal.data,
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend
          ), linetype = "dashed"
        )
    }
    if (!is.null(marginal.data2)) {
      plot <- plot + geom_point(
        data = marginal.data2,
        mapping = aes(
          x = x, y = inv_scale_function(yhat, scale, shift),
          color = marginal.legend2
        )
      ) +
        geom_smooth(
          data = marginal.data2,
          se = FALSE,
          mapping = aes(
            x = x, y = inv_scale_function(yhat, scale, shift),
            color = marginal.legend2
          ), linetype = "dotted"
        )
    }
  }

  subtitle <- glue(
    '<span style="color:{colors["gray"]}">**Absolute frequency (active and delisted)**</span>, ',
    '<span style="color:{colors["blue"]}">**Absolute delisting frequency**</span>, <br>',
    '<span style="color:{colors["orange"]}">**Relative delisting frequency**</span>',
    ifelse(marginal.legend == "", "", paste0(', <span style="color:{colors["magenta"]}">**', marginal.legend, "**</span>")),
    ifelse(marginal.legend2 == "", "", paste0(',<br><span style="color:{colors["lightblue"]}">**', marginal.legend2, "**</span>"))
  )

  plot <- plot + scale_y_continuous(
    name = "Frequency",
    limits = c(min_y_first, max_y_first),
    sec.axis = sec_axis(~ scale_function(., scale, shift),
      name = "Relative delisting frequency"
    ),
    oob = scales::rescale_none
  ) +
  scale_color_manual(
    breaks = c(
      "Absolute frequency", "Absolute delisting frequency",
      "Relative delisting frequency", marginal.legend, marginal.legend2
    ),
    values = c(
      as.character(colors["gray"]),
      as.character(colors["blue"]),
      as.character(colors["orange"]),
      as.character(colors["magenta"]),
      as.character(colors["lightblue"])
    )
  ) +
  scale_fill_manual(
    values = c(
      "Absolute frequency" = as.character(colors["gray"]),
      "Absolute delisting frequency" = as.character(colors["blue"]),
      "Relative delisting frequency" = as.character(colors["orange"]),
      as.character(colors["magenta"]),
      as.character(colors["lightblue"])
    )
  ) +
  labs(
    title = paste0("Histogram of ", feature_description),
    subtitle = subtitle,
    x = feature_identifier
  ) +
  theme(
    plot.subtitle = element_markdown(),
    axis.title.y = element_markdown(color = colors["gray"]),
    axis.title.y.right = element_markdown(color = colors["orange"]),
    legend.position = "none"
  ) +
  geom_text(
    aes(
      x = max_x, y = max_y_first,
      label = "outliers are winsorized"
    ),
    stat = "unique",
    hjust = 1,
    size = 3
  )

  return(plot)
}

# winsorize features
com.winsorized <- data.prepare.winsorize(com, com.features.fin$identifier,
  lower_quantile_boundary = 0.05,
  upper_quantile_boundary = 0.95
)

# plot histogram of features and delisting frequency
plots <- list()
for (i in 1:length(com.features.fin$identifier)) {
  feature_description <- com.features.fin[[i, 1]]
  feature_identifier <- com.features.fin[[i, 2]]

  plots[[feature_identifier]] <- hist_delisting_frequency(
    com.winsorized %>% select(all_of(feature_identifier), years_to_delisting),
    feature_identifier,
    feature_description
  )
  print(plots[[feature_identifier]])
}
```

* Capitalization and Solvency ratios - **capital_ratio, de_ratio, debt_assets, debt_at**: The higher the ratio of debt to assets or liabilities, the higher the relative delisting frequency in the subsequent year.

* Efficiency - **at_turn, inv_turn, rect_turn, pay_turn**: These efficiency ratios but *pay_turn* seem to have no observable marginal effect on delisting. However, if the ratio between cost of goods sold plus the change in inventories and accounts payable gets larger, the observed delisting frequency for the subsequent year decreases.

* Liquidity - **cash_ratio, curr_ratio, quick_ratio**: The ratio between cash plus short-term investments (*cash_ratio*) seems to have no marginal predictive power for delisting. However, if the ratioto current liabilities (*curr_ratio*, *quick_ratio*) increases, the relative frequency of delistings in the subsequent year decreases

* Profitability - **aftret_eq, aftret_equity, gpm, opmad, opmbd, roa, roe**: If these profitability ratios are negative, the observed relative frequency for delisting in the subsequent year is relatively high compared to the cases in which the ratios are positive. The denominators of these profitability ratios contain assets or sales key figures which are greater than or equal to zero. So if the income or sales key figures in the numerator of the ratios are negative, the company's situation is on average more problematic with respect to delisting.

* Valuation - **pe_exi**: For positive price-to-earnings ratios (per share) the observed delisting frequency in the subsequent year becomes smaller for increasing ratios. The higher the price-to-earnings ratio, the better is a share valued. If the price-to-earnings ratio is negative, then the earnings in the denominator of the ratio are negative. If the loss per share is relatively small, then the price-to-earnings ratio is ≪ −1. Hence, decreasing ratios are associated with a lower relative delisting frequency.

```{r, features_distribution_delisting_frequency.plot}
(plots[["capital_ratio"]] | plots[["curr_ratio"]]) / (plots[["roe"]] | plots[["pe_exi"]])
```
